---
title: "Rulia: julia for R"
#author: "RCqls"
output: github_document
---
<!-- Rscript -e "rmarkdown::render('README.Rmd')";rm README.html -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started

This is an attempt to embed the `julia` language in `R`. Actually, very
basic `julia` types are converted to `R` objects.

## Install

1.  Install Julia (all Operating System)

Install [Julia](https://julialang.org/downloads/). For Windows users
don’t forget to select `PATH` in the installer.

2.  Windows user setup

- Install [RTools](https://cran.r-project.org/bin/windows/Rtools/) and a
  terminal with bash (for instance, [Git
  Bash](https://gitforwindows.org/))
- Add Rscript in the `PATH` environment variable (see for example [this
  page](https://www.hanss.info/sebastian/post/rtools-path/))

3.  Bash installation (all Operating Systems)

In a terminal (tested on macOS M1 with julia-1.9.2:) with `julia` and
`Rscript` binaries supposed to be in the `PATH` environment variable,

``` bash
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/rcqls/Rulia/HEAD/inst/install.sh)"
```

## Quick live session

```{r}
require(Rulia)
jl(`1`)
jl(1)
v_jl <- jl(c(1,3,2))
v_jl
R(v_jl)
```

# How it works for the user: the `jl()` function

1. `Rulia` package when loaded, initializes a `julia` session useable inside the current `R` session.
1. `jl()` is the only user friendly function to use in order to:

* execute regular  `julia` code inside `R`
* convert `R` object in `julia` object
* call `julia` function returned by `jl()` function itself
* define `julia` variable directly inside the `julia` session

## `jl()` as evaluation of `julia` expressions

Thanks to the `jl()` function, `Rulia` allows us to call `julia` (possibly multilines) expression given with expression between backtip "\`" (i.e. of class `name` or type `symbol`).

```{r}
jl(`[1,3,2]`)
jl(`[1.0,3.0,2.0]`)
jl(`(a=1,b=[1,3])`)
jl(`[
    1.0,
    3.0,
    2.0
    ]`)
```

All these commands return `jlvalue` objects which are `R` external pointers
wrapping `jl_value_t*` values.

## `jl()` as `julia` converter of `R` objects

A lot of `R` objects can be converted in `julia` objects by simply put them as argument of the `jl()` function.

```{r}
require(Rulia)
jl(c("one","three","two"))
jl(c(TRUE,FALSE,TRUE))
jl(c(1L,3L,2L))
jl(TRUE)
jl(1L)
jl(1)
jl("1")
jl(matrix("one"))
jl(list(a=c(TRUE,FALSE,TRUE), b=1L))
```

## Rulia in the statistic context

* `DataFrame` (`julia` side) and `data.frame` (`R` side)

```{r}
require(Rulia)
jlusing(DataFrames)
jl(`(a=1,b=DataFrame(a=1:3,b=2:4))`) -> nt_jl
nt_jl
list(jltypeof(nt_jl), typeof(nt_jl), class(nt_jl))
```

As expected, `Rulia` offers conversion in both directions

```{r}
nt_R <- R(nt_jl)
nt_R
```


```{r}
jl(nt_R)
```

* `CategoricalArray` (`julia` side) and `factor` (`R` side)

```{r}
require(Rulia)
jlusing(CategoricalArrays)
ca_jl <- jl(`categorical(["titi","toto","titi"])`)
ca_jl
list(jltypeof(ca_jl), typeof(ca_jl), class(ca_jl))
```

```{r}
ca_R <- R(ca_jl)
ca_R
```

```{r}
jl(ca_R)
```

# Rulia in low level mode

# R Finalizers

Following the documentation on embedding `julia`, a system of preserved
references to `julia` values has been created. An `R` finalizer is
assiocated to each `jlvalue` object (in fact, an `R` external pointer
wrapping some `jl_value_t*` value). Whenever the `jlvalue` is gabarged
collected, the reference on the associated `julia` value is also
dereferenced which is then cleaned up by the `julia` garbage collector.
