---
title: "`Rulia`: `julia` for `R`"
#author: "RCqls"
output: github_document
---
<!-- Rscript -e "rmarkdown::render('README.Rmd')";rm README.html -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`Rulia` is an attempt to provide an `R` package in order to facilitate the creation of `R` package for "wrapping" `julia` package. It can also be viewed as a tool similar to `Rcpp` but using the `julia` language instead of `C++`. 

Also, you can visit to [Rencontres R 2024 (Vannes)](https://cqls.dyndoc.fr/talk/Rulia) to have a presentation of the `Rulia` package.

<details>
  <summary><h1>Getting started</h1></summary>

This is an attempt to embed the `julia` language in `R`.

Let us notice that there already exist alternatives `R` packages (see [`JuliaCall`](https://github.com/Non-Contradiction/JuliaCall) README page for a complete list).
[`JuliaCall`](https://github.com/Non-Contradiction/JuliaCall) is the main one. However, the big difference between `Rulia` and `JuliaCall` is that `JuliaCall` depends on the `R` package `Rcpp` and the `julia` package `RCall.jl`.
In other words, `Rulia` only depends on the C APIs of `R` and `julia`. There is then **no dependencies** (except `julia`). 

Also `Rulia` is the next step of the preliminary project called [`jl4R`](https://github.com/rcqls/jl4R) started more than 10 years ago. The author thinks that `Rulia` is a more funny name than `jl4R`. 

## Install

1.  `julia` installation (all Operating Systems)

    Go to [Julia](https://julialang.org/downloads/)

    1. For any Operation system (Windows, MacOS and linux), as proposed first in the `julia` download page, prefer the `juliaup` installation one.
    It offers multi-installation of different versions of `julia`.
    2. Alternative installation (**to avoid** if possible)is to install `julia` from a binary installer to download. 
    For Windows users don’t forget to select `PATH` in the installer

2.  Install `Rulia`

    * From binary (**Windows user only**)

        1. Donwload [Rulia-0.0.1.zip](https://github.com/rcqls/Rulia/releases/download/v0.0.1/Rulia_0.0.1.zip) and install it inisde R
        2. Inside a terminal:
        Whether `julia` is installed with `juliaup` or you specified the `PATH` when launching the binary installer, loading (`library(Rulia)/require(Rulia)`) `Rulia` inside `R` would normally just work.
            

    * From source (**all Operating System**)
    
        1. Windows user need the following setup:

            - Install [RTools](https://cran.r-project.org/bin/windows/Rtools/) and a
            terminal with bash (for instance, [Git
            Bash](https://gitforwindows.org/))
            - Add `Rscript` in the `PATH` environment variable (see for example [this
            page](https://www.hanss.info/sebastian/post/rtools-path/))
        2.  You need the `remotes` R package. 
        3.  Bash installation (all Operating Systems): click the `copy` button to copy the following line and paste in in a `bash` terminal:

``` bash
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/rcqls/Rulia/HEAD/inst/install.sh)"
```

3. Install the followiwng `julia` packages required for `Rulia` in statistic mode: `DataFrames`, `CategoricalArrays`.

## Quick live session

```{r}
require(Rulia)
jl(`1`)
jl(1)
v_jl <- jl(c(1,3,2))
v_jl        # notice the julia output format 
class(v_jl)
typeof(v_jl)
jltypeof(v_jl)
R(v_jl)     # here the R output format 
## a potentially useful task is to call a julia fonction applied on an R ao object
jl(sum)(c(1,3,2))           # the result is a julia object (here a jlvalue R object)
# and then get the result as an R object
jl(sum)(c(1,3,2)) |> R()    # corresponding in the julia side to `sum([1.0, 3.0, 2.0])`
```

The only thing to do in order to initialize `julia` is to load the library `Rulia`.

Then, it is pretty direct to:

1. convert an `R` object to `julia` object (in fact, a `jlvalue` external pointer in the `R` side)
1. apply a `julia` function to the `R` object
1. and finally convert the `julia` result to an `R` object
</details>

<details>
<summary><h1><code>Rulia</code> in user-friendly mode</h1></summary>

1. `Rulia` package when loaded, initializes a `julia` session useable inside the current `R` session.
1. `jl()` is the **only user friendly function** to use in order to:

* execute regular  `julia` code inside `R`
* convert `R` object in `julia` object
* call `julia` function returned by `jl()` function itself
* define `julia` variable(s) directly inside the `julia` session

<details><summary><h2><code>jl()</code>: <code>julia</code> code evaluation</h2></summary>

Thanks to the `jl()` function, `Rulia` allows us to execute `julia` (possibly multilines) expression given with expression between backticks "\`" (i.e. of class `name` or type `symbol` in the `R` side).

```{r}
jl(`[1,3,2]`)
jl(`[1.0,3.0,2.0]`)
jl(`(a=1,b=[1,3])`)
jl(`[
    1.0,
    3.0,
    2.0
    ]`)
```

All these commands return `jlvalue` objects which are `R` external pointers
wrapping `jl_value_t*` values.
</details>
<details><summary><h2><code>jl()</code>: <code>julia</code> converter of <code>R</code> objects</h2></summary>

A lot of `R` objects can be converted in `julia` objects by simply put them as argument of the `jl()` function.

```{r}
jl(c("one","three","two"))
jl(c(TRUE,FALSE,TRUE))
jl(c(1L,3L,2L))
jl(TRUE)
jl(1L)
jl(1)
jl("1")
jl(matrix("one"))
jl(list(a=c(TRUE,FALSE,TRUE), b=1L))
jl(2 * sin(1:3))    # this is a R call
2 * sin(1:3)
```
</details>
<details><summary><h2><code>jl()</code>: <code>julia</code> function  call inside <code>R</code></h2></summary>

The main use of the `Rulia` package is to call `julia` function (in fact, `julia` method because of the implicit **multiple dispatching** provided by `julia`) inside the `R` system. The more challenging goal of `Rulia` is to try to provide a `R` syntax to call `julia` function which as most as possible close to the original `julia` syntax.

Let us start with a simple example.

```{r}
## An utility function to fix the seed of Random number in julia

```

```{r}
jl_set.seed(12) # to fix the seed 
jl(rand)(`2`)   # julia integer
jl(rand)(2L)    # implicitly converted R integer
```

Also `jl_set.seed()` is a facility function equivalent to:

```{r}
jl_set.seed
jlusing(Random)
jl(`Random.seed!`)(12L)
```

In fact both these lines are user-friendy simplified versions of what would be necessary to call:

```{r}
jl(rand)(jl(`2`))   # julia integer
jl(rand)(jl(2L))    # implicitly converted R integer
```

The challenging primary goal in `Rulia` is:

    An expression in `Rulia` only need a unique `jl()` call (whenever many `jl()` calls would be normally necessary).

How is a such trick possible?

Let us first observe the result returned  when the argument of `jl()` is an expression of a `julia function`.

```{r}
jl(`sum`)       # the usual way
jl(sum)         # which is equivalent to the simplified way thanks to R
class(jl(sum))  # this is not directly a jlvalue R object
```

Let us comment what is special here. `jl(sum)` should normally returns an `R` object of class `jlvalue`.
But since our goal is to apply the function, `jl(sum)` is tranformed in a `jlfunction` that can be called with arguments that need to be `R` objects of class `jlvalue`.
Thanks to the metaprogramming provided by `R`, one only needs to provide the arguments of the `jlfunction` with:

* `R` objects implicitly converted to `jlvalue` objects  
* `julia` expressions given between backticks also implicitly executed (for you) in the `julia` side to finally provide `jlvalue` results  

The main point is that no need of `jl()` is required whe specifying arguments of the `jlfunction`.

Notice also that the `rand` `julia` function needs an integer as argument so:

```{r}
jl(rand)(2)    # fails (use summary R generic function to have the complete julia output)
```

`julia` function with keyword-arguments can be called too:
```{r}
jl(sum)(1:10)           # an integer
jl(sum)(1:10, init=12)  # a double
```

</details>

<details><summary><h2><code>jl()</code>: <code>julia</code> variable(s) from <code>R</code></h2></summary>
```{r}
jl(a=jl(rand)(2L), b=1:3)
jl(a)
jl(b)
```

A special conflict case may happen now if `b` is also a `R` variable.

```{r}
jl(b)
b <- 10
jl(b)
## Also notice that
jl(`b`) # Not a julia variable since jl(`b`) is equivalent to jl(b) in R
## To access the b julia variable
jl()$b  # as explained in the next section
```
</details>

<details><summary><h2><code>jl()</code>: <code>julia</code> variables environment</code></h2></summary>

Without any argument, `jl()` returns the list of all `julia` variables in the `Main` module.

```{r}
jl()
```

It is also possible to access a specific `julia` variable from the `julia` variables environment `R` object. 

```{r}
jl()$b  # b variable in Main module
jl()$c  # c does not exist and then fails
```

</details>

<details><summary><h2>
<code>R()</code>: <code>R</code> converter of <code>julia</code> objects
</h2></summary>

The converse conversion of `jl()` is `R()` 

```{r}
R(jl(rand)(2L))
jl(rand)(2L) |> R()
```
 </details>

<details><summary><h2><code>Rulia</code> in the statistic context</h2></summary>

* `DataFrame` (`julia` side) and `data.frame` (`R` side)

```{r}
require(Rulia)
jlusing(DataFrames)
jl(`(a=1,b=DataFrame(a=1:3,b=2:4))`) -> nt_jl
nt_jl
list(jltypeof(nt_jl), typeof(nt_jl), class(nt_jl))
nt_jl$b # suer-friendly access of a julia NamedTuple in the R style
```

To compute `julia` code needs to be put between two backticks and not between quote or double quote (which is a regular `R` character object to be converted in `julia`). It is better to insist, don't confuse the third line before and the first following one (which returns a simple `julia` object of type `String`):

```{r}
jl("(a=1,b=DataFrame(a=1:3,b=2:4))") -> str_jl
str_jl
list(jltypeof(str_jl), typeof(str_jl), class(str_jl))
```

As expected, `Rulia` offers conversion in both directions, `julia` to `R` and conversely `R` to `julia`

```{r}
nt_R <- R(nt_jl)
nt_R
```

and conversely `R` to `julia`

```{r}
jl(nt_R)
jl(nt_R$b)
```

* `CategoricalArray` (`julia` side) and `factor` (`R` side)

```{r}
require(Rulia)
jlusing(CategoricalArrays)
ca_jl <- jl(`categorical(["titi","toto","titi"])`)
ca_jl
list(jltypeof(ca_jl), typeof(ca_jl), class(ca_jl))
```

Below, the conversion `julia` to `R`

```{r}
ca_R <- R(ca_jl)
ca_R
```

and conversely, the conversion `R` to `julia`

```{r}
jl(ca_R)
```
</details>
</details>


<details>
<summary><h1><code>UnsafeArray</code> thanks to <code>RCall.jl</code></h1></summary>
Conversion of `R` object to `julia` system can be magically avoided thanks to `RCall.jl`. 
After installing `RCall.jl` and loading `jlinclude(Rulia::RCall)`, one can have access to this feature.
`R()` which is usually used for conversion of `julia` object to `R` object is here exceptionnaly used as a "wrapper" of `R` vector into a `jlvalue` object pointing to a `julia` of type `Array` and `R` class `UnsafeArray` (since derived from the `unsafe_array()` `julia` function introduced by `RCall.jl`) sharing the same memory of the original `R` vector.
This feature as illustrated below can be applied to `R` vector of type `double`, `integer`, `complex` but not `character`. Notice that `logical` vector is considered in `julia` as a `Vector{Int32}` since it is the natural representation of `logical` in `R`.

```{r}
jlinclude(Rulia::RCall)
zz <- runif(3)
zz
Rzz <- R(zz) # this is a jlvalue object wrapping zz
Rzz
class(Rzz)
jl(typeof)(Rzz)
Rzz[1] <- 2
Rzz
## and magically (no conversion)
zz
```

`Rzz` is the viewed in the `julia` side as a true `Vector{Float64}` pointing exactly to address of the `zz` vector.  
Modifying `Rzz` directly modifies `zz`.

These features also apply for `factor` (the `levels` part being copied in the `julia` side) and for `data.frame` (containing exclusively variables accepting this "wrapping" mode).

```{r}
jlinclude(Rulia::RCall)
fa <- factor(c("toto", "titi", "toto"))
fa
Rfa <- R(fa) # this is a jlvalue object wrapping fa
Rfa
class(Rfa)
jl(typeof)(Rfa)
Rfa[1] <- "titi"
Rfa
## and magically (no conversion)
fa
```

The main use of this feature is rarely to define `Rzz` and `Rfa` but to directly use `R(zz)` and `R(fa)` as argument(s) of a `julia` function.  

```{r}
jl(`
function f(x)
    x .= x .+ 2 
end
`)
 
jl(f)(R(zz))
## and the magic part
zz
```

Important to notice that no change of dimension has to be done in the `julia` side. The `julia` wrapper can only read or update value(s).

</details>

<details>
<summary><h1>More details on <code>jl()</code></h1></summary>
TODO
</details>

<details>
<summary><h1><code>Rulia</code> in low level mode</h1></summary>

<details><summary><h2>safe <code>jleval</code> mode</h2></summary>

In `Rulia`, `jl` mode offers a way to call a **safe** low level mode called `jleval` mode that relies mainly to three main functions: 

* `jlvalue()` to convert `R` object to `jlvalue` wrapper of a `julia` object (as already seen previously)
* `jleval()` to evaluate a `julia` expression as its character argument
* `jlcall()` to call function by its name given as a character and safe since protected by a try/catch `` 

In fact, `jl` mode uses the metaprogramming and lazziness offered by `R` to avoid the use of quote in order to write `julia` code as expressed in the foolowing example.

```{r}
jleval("[1,3,4]")   # jl(`[1,2,3]`)
jleval("VERSION")   # jl(VERSION)

jleval("            
f(x,y) = x + y
(f(2,3), f(1.0,3))
")

## jlvalue() is faster than jl() here
jlvalue(TRUE)       # jl(TRUE)
jlvalue(1L)         # jl(1L)
jlvalue(1)          # jl(1)
jlvalue("1.0")      # jl("1.0") 
jlvalue(c(TRUE, 1L, 1, "1.0"))      # jl(c(TRUE, 1L, 1, "1.0"))
jlvalue(list(TRUE, 1L, 1, "1.0"))   # jl(list(TRUE, 1L, 1, "1.0"))

jleval('a =[true, 1, 1.0, "1.0"]')  # jl(`a =[true, 1, 1.0, "1.0"]`)
jleval('a')                         # jl(a)
jleval('b = (true, 1, 1.0, "1.0")') # jl(`b = (true, 1, 1.0, "1.0")`)
jleval('b')                         # jl(b)
## error below don't crash
jleval('b = (true, 1, 1.0, "1.0"')  # jl(`b = (true, 1, 1.0, "1.0"`)

jleval("sum")                       # jl(sum)
jleval("typeof(sum)")               # jl(typeof)(sum)
jlcall("sum", jleval("[1,3,2]"))    # jl(sum)(`[1,3,2]`)
jlcall("sum", c(1, 3, 2), init = 4) # jl(sum)(c(1,3,2), init=4)
jlcall("isa", jleval("sum"), jleval("Function"))    # jl(isa)(sum, Function)
jlfunc(jleval("sum"), c(1,3,2), init = 4)           # in fact it is what jl(isa) does
jleval("sum isa Function")                          # jl(`sum isa Function`)
```

The bad part of this safe low level mode is the performance issue. Indeed, these functions are not the most efficient since they are not as closed as the `julia C API`. An unsafe low level mode, called `jlvalue_eval` mode, naturally exists in `Rulia` that express the closest as possible the `julia C API`.
</details>

<details><summary><h2>unsafe <code>jlvalue_eval</code> mode</h2></summary>

As expressed before, this mode is **unsafe** and the user should be sure that the `julia` expression is correct. 
One can think of using this mode in some development package where efficiency really matters.

`jleval()` and `jlcall()` functions are then replaced by `jlvalue_eval()` and `jlvalue_call()` functions respectively.

```{r}
jlvalue_eval("[1,3,4]")
jlvalue_eval("VERSION")

jlvalue_eval("
f(x,y) = x + y
(f(2,3), f(1.0,3))
")

jlvalue_eval('a = [true, 1, 1.0, "1.0"]')
jlvalue_eval('a')
jlvalue_eval('b = (true, 1, 1.0, "1.0")')
jlvalue_eval('b')
## error below would crash badly
# jlvalue_eval('b = (true, 1, 1.0, "1.0"')

jlvalue_eval("sum")
jlvalue_eval("typeof(sum)")
jlvalue_call("sum",jlvalue_eval("[1,3,2]"))
jlvalue_func(jlvalue_eval("sum"),jlvalue_eval("[1,3,2]"))
## Notice that this is not possible: jlvalue_call("sum", jlvalue([1,3,2]), init=4)")
```

Also, in this mode

</details>

</details>

<details>
<summary><h1>Pure <code>R</code> package made with <code>Rulia</code></h1></summary>
TODO
</details>

<details>
<summary><h1><code>Rulia</code> utility functions</h1></summary>
TODO
</details>

<details>
<summary><h1><code>R</code> Finalizers</h1></summary>

Following the documentation on embedding `julia`, a system of preserved
references to `julia` values has been created. An `R` finalizer is
assiocated to each `jlvalue` object (in fact, an `R` external pointer
wrapping some `jl_value_t*` value). Whenever the `jlvalue` is gabarge
collected, the reference on the associated `julia` value is also
dereferenced which is then cleaned up by the `julia` garbage collector.

Since the `julia` session is  not persistent when the `R` session is, what happens if a `jlvalue` object is still in the Workspace (environment return by `globalenv()` or `.GlobalEnv`).

</details>

<details>
<summary><h1>Benchmark <code>Rcpp</code> vs <code>Rulia</code></h1></summary>
```{r load, include=FALSE}
require(microbenchmark)
load("../microbenchmark/RcppVsRulia.RData")
options(width=120)
```

```{r}
summary_mbs <- round(sapply(1:6, function(i) sapply(1:28, function(k) mean((mbs[[i]][mbs[[i]]$expr == levels(mbs[[i]]$expr)[3],]$time)) / mean((mbs[[i]][mbs[[i]]$expr == levels(mbs[[i]]$expr)[k],]$time)))),2)
rownames(summary_mbs) <- levels(mbs[[1]]$expr)
colnames(summary_mbs)<- names(mbs)
summary_mbs
```

```{r}
mbs[["n=1000"]]
```

```{r}
mbs[["n=10000"]]
```

```{r}
mbs[["n=1e+05"]]
```

```{r}
mbs[["n=1e+06"]]
```

```{r}
mbs[["n=1e+07"]]
```

```{r}
mbs[["n=1e+08"]]
```

</details>